Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-05-20 doc/response.tex

Thank you for the chance to revise and resubmit.
We apologize for the delay.
Reviewer 1 suggested that we should divide the paper into two: theory and application.
Given the relevance of this manuscript to the current COVID-19 outbreak, we think it is best to write an application-based one first (this one) and then write a theory-based on later.
We believe that incorporating some theory with application will help readers understand the material better at the moment.
Otherwise, we have tried our best to incorporate the reviewers' suggestions.
See below for detailed response.

\rev{Reviewer \#1}

\revtext{The main objective of this paper is to correct selection bias in observational
(contact tracing) data to yield unbiased estimate of the "true" distribution
corresponding to generation intervals defined in Lines 42-43 (if I understand
properly).}

Lines 42-43 refers to a single measurement, rather than a distribution. We have made this clear (see next comment).

@@

\revtext{Lines 42-43: "A generation interval is defined as the time between when a
person becomes infected and when that person infects another person."
(with reference [10]). Indeed, most of the cited literature in this paper
start with the definition exactly the same way. I feel that this definition
is imprecise and is not as obvious as it reads, especially in its relationship
with the intrinsic generation interval (Line 46) and its distribution (Lines
49,101). It is not proven that they are equivalent.}

We make clear that this definition corresponds to the definition at the individual-level (i.e., a single interval), which is unambiguous, and distinguish it from the population-level distributions:

``At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.''

@@

\revtext{Line 46: "...intrinsic generation intervals..." This is not yet defined. Is it
equivalent to the definition in Lines 42-43?}

We now define intrinsic generation interval properly:

``The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.''

@@

\revtext{Line 46: "......intrinsic generation intervals measure the infectiousness of an
infected individual." This statement is vague. I will get back to this phrase
a few pages later in the discussion on "equilibrium distributions" and
suggest ways to make it more precise.}

We have now made this definition more precise. In Section 2, we also introduce the idea that the intrinsic distributions can be thought of as equilibrium distributions:

``$g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
Since all infectious contacts lead to infection in a disease-free equilibrium of a homogeneously mixing population, $g(\tau)$ also describes the expected time distribution of secondary cases caused by an average primary case in a fully susceptible population and can be treated as an ``equilibrium'' distribution of the corresponding homogeneous mixing model.''

@@

\revtext{Lines 49,101: "...the intrinsic generation interval distribution..." While the
interval itself is not defined in Line 46 (in my opinion), its distribution is
clarified in Line 101 in relation to equation (2) under homogeneous mixing.
I hereby pose the following questions:}

\revtext{
1. Although $g(\tau) > 0$ satisfies all the properties of a probability density
function (p.d.f.) of a non-negative continues random variable, does that
automatically correspond to a specific random variable relevant to any of
the observable (or conceptually observable) "generation intervals"?
}

We now define $g(\tau)$ as follows: ``expected time distribution of infectious contacts by a primary case''.

@@

\revtext{
2. Is it the p.d.f. corresponding to the definition in Lines 42-43? Proof?
}

Definition in Lines 42-43 does not correspond to the definition of the intrinsic generation-interval distribution. 
It is a single measurement --- we tried to make it clear that it is not directly connected to any p.d.f.
We have tried to make this clearer in the text.

\revtext{
Lines 44-48: "... the observed generation-interval distribution can change...
...observed generation intervals refer to the time between actual infection
events." The actual infection events correspond to the exact moments of
passing the infection between a pair of infected-susceptible individuals.
Are they observable? In most cases, including contact tracing, observable
events are clinical (e.g. onset of symptoms). The relationship between
the interval defined by the time between actual infection events and the
interval defined by the time between clinical onsets is unclear. At least the
latter depends on the incubation period (of the infectee), in addition to the
latent period and the infectious period (of the infector). The incubation
period is defined as the time from infection to symptom onset, and is
different from the latent period such as in the SEIR model. This paper
does not make such distinction. It assumes that the time between actual
infection events "were observable" in a hypothetical way.
}

@@

\revtext{
Suggestion: Re-phrase "observed generation intervals" or make some further
clarification on this. This is important because naive readers may tend
to take the face value and collect different kinds of observed data and
make inferences, such as that based on clinical events. This is potentially
misleading.
}

We agree that calling it ``observed'' can be misleading. We now refer to them as \emph{realized} intervals.

``There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time between infection of a primary case and infection of its secondary cases in a fully susceptible population,
while \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Due to individual variation in infection time, the realized generation-interval distribution (at the population-level) can change depending on when and how it is defined \citep{svensson2007note, kenah2008generation, nishiura2010time, champredon2015intrinsic}.''

We have also added the following sentence as per suggestions in previous comments:

``While realized generation intervals often cannot be observed in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
we typically want to try to make inference based on all available data.''

@@

\revtext{Lines 54-55: ``the backward generation interval...'' This is a key measurement in this paper along with the length-bias in relation to data observation scheme. Conceptually, there may also be a "forward" generation
interval. Both concepts were initially raised in Svensson (2007, ref.[21]
in the paper), who pointed out the difference in distribution between the
primary generation time as measured prospectively from the time of the
infection of the infector to the transmission to the infectee, and the secondary generation time as measured retrospectively from the time of the
transmission to the infectee to the infection of the infector.}

We now introduce the forward distribution as well.

\revtext{1. does that mean that your objective is to make unbiased inference on the
"forward", or the primary, generation time?}

Yes, we want to make inference on the \emph{initial} forward generation-interval distribution, rather than the intrinsic generation-interval distribution.
We have tried to make this clear and no longer use the term ``effective'' generation-interval distribution.

@@

\revtext{2. does the definition in Lines 42-43 correspond to the forward generation
interval?}

No. Lines 42-43 is simply a single measurement and does not correspond to any specific generation interval.

@@

\revtext{4. does the definition in Lines 42-43 correspond to a random variable with
p.d.f. given by $g(\tau)$?}

No. See above.

\revtext{5. do you imply that the intrinsic generation interval (not defined in Line 46
but distributed according to Line 101) and the forward measure, and the
objective of the unbiased inference, are all equivalent?}

No. The intrinsic distributions does not account for any spatial effect and therefore are different from the forward distributions.
We have tried to make this clear throughout the manuscript.
In particular, we provide the following sentence in Introduction to distinguish intrinsic from the realized distributions:

``Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.''

\revtext{
The "the backward generation interval (or the secondary generation time)" is statistically sound. An infectee, before being infected, has been exposed to many infectors in a prevalent cohort and
could be independently infected by any of them. Each infectee is only associated with a single backward generation time whose distribution arises
from an independent competing risk framework. At the population level,
all backward measured intervals are independent random variables (but
not necessarily identically distributed). On the other hand, the "forward
generation interval" is problematic. A typical infected individual may
infect several contacts, but not independently if the infectious period is
random. Therefore, it is not appropriate to consider that the times to
transmission of the pathogen to its infectees as an iid. sample. This can
be shown in Exercise 3.5 (f) of Yan and Chowell (2019). At the population level, should such data can be observed, there might exist correlated
intervals arising from the same infector, something to ponder about.
}

\swp{I don't know.}

\revtext{This paper addresses length-bias in the "backward generation interval" and
different ways of adjusting this bias, assuming the intervals were observable
especially in the contact tracing setting. This is scientifically important, not
only under the context of the current paper, but also applicable in many other
contexts in science. Several issues need to be discussed.}

Thank you. We have tried to address the issues outlined in the review.

\revtext{
Line 43: "...individual variation in infection time..." I presume the authors
meant individuals infected at different (calendar) times in the course of
an outbreak. If this is the case, I suggest re-wording to avoid confusion
}

We have removed this sentence.

\revtext{
Lines 59, 114-115, 129, etc.: There is a distinction between "truncation"
and "censoring" in survival analysis. In the current context (e.g. contact tracing), when the inclusion criteria is conditioning on the infection
event of the infectee and looking backward to ascertain the infection time
of its infector, it is right-truncation. It leads to over-sampling short durations (as stated in Line 129). I suggest to use "right-truncation" instead
of "censoring" throughout the paper, wherever it appears.
}

We now use truncation instead of censoring.

\revtext{
Lines 121 and 123: The backward generation-interval distribution $b_t(\tau)$ and
the "censored" (right-truncated?) generation-interval distribution $c_t(\tau)$
should be clearly defined and expressed (e.g. in terms of conditional distribution) before expressing the relationship (6).
}

Done.

\revtext{
Lines 233-241: Once more, the sampling bias corresponds to right-truncation,
not censoring. A great deal of words is involved leading to gef $g_{eff}(\tau)$. The
mathematical language is more powerful and perhaps gef $g_{eff}(\tau)$ could be
explained more concisely (and take less space?)
}

We have removed this paragraph and tried to make the explanation more concise.
We no longer use the term effective generation-interval distribution.

\rev{Lines 241-243: "In a large homogeneously mixing population, the e§ective
distribution is expected to be equivalent to the intrinsic distribution..."
Define both $g(\tau)$ and gef $g_{eff}(\tau)$ mathematically at the first place and prove
their equivalence in a large homogeneously mixing population.}

Done.

\rev{However, I suggest the authors to compare the "individual-based
method" in the paragraph starting from Line 252 to Line 260 with the statistical methods for analyzing right-truncated survival data as mentioned
above, and explicitly write out the likelihood function (Line 258) to see if
they are the same or different.}

\rev{The authors has used the notations $\RR$ (Line 35) and $\Ro$ (Lines 98-99) for reproduction numbers and
linked them via $\RR = \Ro S(t)$:}

\rev{Shouldnít $\RR$ be a function of time $t$ via $S(t)$?}

We no longer use $\RR$ to avoid confusion. We now use $\Rini = \Ro S(0)$ to reflect the possibility that $S(0)$ may not be 1.

\rev{Is $S(t)$ a number or a proportion? In deterministic models, this matters
less as one can argue that all deterministic models are describing the
change of proportions (by re-scaling on population size, assuming it is
constant); but it matters a lot in stochastic models because the population
size is an essential parameter.}

Proportion. Made this clear.

\rev{
When stating that ``$\RR$'' transcends to the final size of the epidemic (Line
37), do the authors actually mean $\Ro$; which does not depend on time?
}

Yes.

\rev{
When talking about the final size, one considers that the system is at a
disease-free equilibrium. When using the final size equation to calculate a
value for the reproduction number, the calculated value does not depend
on time. If there are behavior changes and public health interventions
throughout the outbreak, the calculated value $\RR_{final}$ is usually smaller
than $\Ro$ (defined at another disease-free equilibrium at the very moment
when an infector was seeded). So now we have three reproduction numbers, $\Ro$, a time-dependent $\RR = \Ro S(t)$ and $\RR_{final}$: Need to clarify the differences and reveal the hidden assumptions when discussing (and using)
each of them.
}

Discussion!

\rev{
The authors stated that $\RR$ transcends to the final size of the epidemic
(with cited references). I boldly assume that the authors actually meant
by $\Ro$: This statement is in the context of a large outbreak that ends
when either there is no infectious individuals or there is no susceptible
individuals left in the population. Under a different context, the outbreak
may reach an endemic steady state with sustained on-going transmission
persisting for a very long period of time due to re-infection or demography.
It could be further said that $\Ro$ can transcend to the long term endemicity
level, such as $S(\infty) = \RR$: I leave to the authors to search the literature
for the latter
}

Yes, we meant $\Ro$. We have now clarified this in the text.
We now mention in Introduction that $\Ro$ can also affect the endemicity level and added a few appropriate citations.

\rev{$g(\tau)$ as a p.d.f. given by (2) and Line 101 is also an equilibrium
distribution.}

We disagree with the reviewer that the intrinsic generation-interval is an equilibrium distribution.
It is only an equilibrium distribution for a model that assumes a large, homogeneously mixing population.
As we show in this paper, when there is spatial structure, the realized generation-interval distributions differ from the intrinsic one. 
In particular, even when the entire population is susceptible, the forward generation-interval distribution of the single primary infector will still be subject to the egeocentric effect.

\rev{ Under suitable assumptions on homogeneous mixing, the intrinsic
generation interval distribution $g(\tau)$ measures the infectiousness within
an infected individual without involving another individual. (In SEIS,
SEIR, SEIRS settings, it is a function of the latent period distribution
and the infectious period distribution.)}

We agree that $g(\tau)$ measures the infectiousness within an infected individual without involving another individual.
We defined the intrinsic generation-interval distribution as follows:

``The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.''

\revtext{The distribution $g(\tau)$ links the two intrinsic parameters R0 and r
(homogeneous mixing). Both $\Ro$ and $g(\tau)$ are intrinsic in the context of the equilibrium conditions. Therefore the parameter $r$ is a
theoretical value and is very di§erent from any observed exponential
or sub-exponential growth rate as the system is not at equilibrium.}

\rev{Observed exponential or sub-exponential growth events are different from the infection events. The current paper only addresses the infection events (mostly unobservable). This was
already addressed in the comments with respect to Line 47-48.
Any attempt to link the estimated $r$ based on data collected
on observed events using the Euler-Lotka equation could be extremely misleading}

We point out in the Discussion that estimated $r$ can be different from the true $r$. 
However, it is not \emph{extremely} misleading to use the estimated $r$ from the observed data. 
See:

``Ma, Junling, Jonathan Dushoff, Benjamin M. Bolker, and David JD Earn. "Estimating initial epidemic growth rates." Bulletin of mathematical biology 76, no. 1 (2014): 245-260.''

\rev{
The intrinsic parameters R0 and r are parameters in branching
process, which is used to approximate the "initial phase" of the
outbreak and that approximation is poor. By the way, the condition in Lines 107-108 $S(t) = S$ is also inaccurate. The expected exponential growth is under the approximation $S(t)/n \approx 1$ and
$n \to \infty$; where $n$ is the population size. The so-called initial
phase (when the depletion of the susceptible population is negligible) totally ignores the difference between the initial disease free equilibrium and the highly explosive dynamic growth
}

``The expected exponential growth is under the approximation $S(t)/n \approx 1$ and
$n \to \infty$'' 
This is inaccurate.
It is still possible to observe exponential growth when the population is not fully susceptible ($S(t) \approx S(0) \neq 1$) as long as the depletion of the susceptible population is negligible.
To avoid confusion, we now use the term ``initial reproduction number'': $\Rini = \Ro S(0)$.

\rev{
 It is too strong a statement that "the Erlang distributions are
more realistic than the exponentially distributed periods in the standard
SEIR model." The advantage of using the Erlang distributions is that the
system can be analyzed using ordinary differential equations. They are
only suitable for latent and infectious periods with very small variances
}

The exponential distribution is a special case of the Erlang distribution (when the shape parameter is equal to 1).
We disagree the statement that it is ``too strong'' to say that a more general distribution is more realistic than a special case.
To avoid confusion, we now say refer to Erlang distributions as ``more realistic approximations''.

\rev{
The Erlang distribution is not only a subset of the gamma distribution with integer valued shape parameters, but also it is limited to distributions with variances smaller than the corresponding exponential distribution with the same mean value.
}

The Erlang distribution includes the exponential distribution.

\rev{
One can not even claim the gamma distributions are more realistic
without supported by data. For one thing: all gamma distributions
have exponential tails which may be contested in real data.
}

One can at least claim that the gamma distribution is more realistic than its subset, the exponential distribution.

\rev{
It is more convenient to use the Laplace transforms as a tool.
}

The SEIR model is just an example --- the purpose of this paper is not to go through all possible generalizations of the SEIR model, but rather to explore spatial components.
The Laplace tranforms are only more useful for the intrinsic generation-interval distributions. 
The forward generation-interval distribution is analytically intractable and therefore, the Laplace transforms no longer hold.
Given that our goal is to infer the initial forward generation-interval distribution, rather than the intrinsic distribution, we feel that the Laplace transforms are not necessarily helpful for the discussion.
In particular, the reviewer asks us to prove special cases of the Euler-Lotka equation, which has been proven numerous times as suggested by the references, using Laplace transforms; 
we do not think this is any relevant to the discussion of our paper.
We cite all Yan et al papers that were mentioned in the review in our main text.

\rev{
How to assess if the gamma distribution is a good choice as a
model? Run some kind of goodness-of-fit tests? How does the likelihood
function look like? Since the "observed" generation-intervals (assuming
observable) do not arise from a designed random experiment, but observational data, I expect the likelihood function should reflect the observation
scheme (e.g. retrospective)? Please be explicit.
}

\rev{
Elaborate how the estimation of $\RR$ is expected to be sensitive to under-reporting (of which data)
}

\rev{
How exactly is $\RR$ (and, is it \Ro or some kind of time-dependent \RR) estimated in this paper: via the Euler-Lotka equation or using the approach in White and Pagano (2007, ref. [54] in the paper)?
}

``the inferred generation-interval distribution is then used to estimate the reproductive number $\Rini$ from the observed growth rate $r$ using the Euler-Lotka equation.''

\rev{
How close is the "related approach" in White and Pagano
(2007, ref. [54] in the paper) related to the current paper? Did they use the
Euler-Lotka equation at all? Did they define and measure the generation intervals that were consistent with the generation-intervals discussed in
the current paper?
}

\rev{
The authors use a logistic function with a scale parameter $r$,
the carrier capacity parameter $K$ and the initial condition $c_0$ to model the
cumulative incidence $c(t)$: Assuming, (theoretically), the infection events
are observable. It is inaccurate to claim that r is the exponential growth
rate
}

### 2020-05-20 doc/network.tex
working on it...

\newcommand{\eref}[1]{(\ref{eq:#1})}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\Fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\sref}[1]{Sec.~\ref{#1}}
\newcommand{\frange}[2]{Fig.~\ref{fig:#1}--\ref{fig:#2}}
\newcommand{\tref}[1]{Table~\ref{tab:#1}}
\newcommand{\tlab}[1]{\label{tab:#1}}
\newcommand{\seminar}{SE\mbox{$^m$}I\mbox{$^n$}R}

@@

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

@@

\begin{flushleft}{
	\Large
	\textbf\newline{
		Inferring generation-interval distributions from contact-tracing data
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,2,*}
David Champredon\textsuperscript{3,4}
Jonathan Dushoff\textsuperscript{3,5}
\\

@@

Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
Generation intervals are often measured through contact tracing by identifying who infected whom.
We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including censoring (due to observation time), and the effects of susceptible depletion at various spatial scales. 
Early in an epidemic, we expect the variation in the observed generation intervals to be mainly driven by the censoring and the population structure near the source of disease spread; 
therefore, we predict that correcting observed intervals for the effect of temporal censoring but \emph{not} for spatial effects will provide a spatially informed ``effective'' generation-interval distribution, which will correctly link $r$ and $\RR$.
We develop and test statistical methods for temporal corrections of generation intervals,
and confirm our prediction using individual-based simulations on an empirical network.

@@

Infectious disease modeling, generation interval, basic reproduction number, population structure, contact tracing

@@

An epidemic can be described by the exponential growth rate, $r$, and the reproductive number, \RR.
The reproductive number, defined as the average number of secondary cases arising from a primary case, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition}.
However, estimating the reproductive number directly from disease natural history requires detailed knowledge which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
Instead, the reproductive number can be \emph{indirectly} estimated from the exponential growth rate, which can be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, svensson2007note, roberts2007model, wallinga2007generation, park2019practical}.

At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
While this definition is widely used in the literature, it is not connected directly to a population-level distribution.
There are important distinctions to be made when defining generation-interval \emph{distributions} at the population-level: \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case,
while \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.

While the shape of intrinsic generation-interval distribution is often assumed to be invariant across time, the shape of the realized generation-interval distribution depends on when and how intervals are measured \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual. 
Thus, ``backward'' generation intervals (which look at a cohort of infectees, and ask when their infectors were infected) will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
Likewise, we can define ``forward'' generation intervals, which look at a cohort of infectors and ask when their infectees were infected.
Forward generation intervals can contract over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.

Forward generation intervals are also affected by spatial structure.
In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population. 
This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
As a result, forward generation intervals will have shorter mean than the intrinsic generation-interval distribution.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

In practice, realized generation intervals are often measured by aggregating available information from contact tracing.
While an epidemic is ongoing, these measurements are ``censored'': we don't know what happens after the observation time.
These censored intervals are similar to backward intervals: there is a bias towards observing shorter intervals, which are more likely to have concluded by the observation time.
These observations allow us to make the following prediction: removing the time-censoring bias from observed generation intervals will yield a spatially corrected ``effective'' distribution that contains information about the population structure and will allow us to correctly infer the reproductive number from the exponential growth rate.
We further predict the effective generation-interval distribution to be equivalent to the \emph{initial} forward generation-interval distribution, which describes the renewal process of infection during the exponential growth rate.

In this study, we explore spatiotemporal variation in generation intervals measured through contact tracing.
We extend previous frameworks to study how the censored generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
We classify spatial effects on generation intervals into three levels (egocentric, local, and global) and discuss how they affect observed generation-interval distributions.
Finally, we compare two methods for accounting for temporal bias and test our prediction using simulations.

@@

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components (see \fref{indpop}).
An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``effective contacts'' (contacts which will cause infection if the contactee is susceptible).
Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
represent time since infection.

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations:
\begin{equation}
K(\tau) = \int k (\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
The population-level kernel describes the rate at which secondary infections are expected to be caused by an an infected individual, on \emph{average}.

Assuming that a population mixes homogeneously, we can write: 
\begin{equation}
K(\tau) = \RR_0 g(\tau),
\end{equation}
where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population), 
and $g(\tau)$ is the expected time distribution of those cases (the intrinsic generation-interval distribution).

In a homogeneously mixing population, current disease incidence at time $t$, $i(t)$, is the product of the current infectiousness of individuals infected in the past and the current proportion of the population susceptible, $S(t)$.
\begin{equation}
i(t) = S(t) \int K(s) i(t-s) ds = \RR_0 S(t) \int g(s) i(t-s) ds.
\end{equation}
This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
Over a period of time where the susceptible proportion $S(t) = S$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between speed and strength of an epidemic:
\begin{equation}
\frac{1}{\RR} = \int g(s) \exp(-r s) ds,
\end{equation}
where $\RR = S(t) \RR_0$ is the effective reproductive number.

\section{Generation-interval distributions across time}

>>>>

\newcommand{\eref}[1]{Eq.~\ref{eq:#1}}
\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\Fref}[1]{Fig.~\ref{fig:#1}}
\newcommand{\sref}[1]{Sec.~\ref{#1}}
\newcommand{\frange}[2]{Fig.~\ref{fig:#1}--\ref{fig:#2}}
\newcommand{\tref}[1]{Table~\ref{tab:#1}}
\newcommand{\tlab}[1]{\label{tab:#1}}
\newcommand{\seminar}{SE\mbox{$^m$}I\mbox{$^n$}R}

@@

\usepackage{xspace}

\newcommand{\Rx}[1]{\ensuremath{{\mathcal R}_{#1}}\xspace} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\RR}{\ensuremath{{\mathcal R}}}
\newcommand{\Rini}{\Rx{\textrm{\tiny initial}}}
\newcommand{\Rhat}{\ensuremath{{\hat\RR}}}
\newcommand{\tsub}[2]{#1_{{\textrm{\tiny #2}}}}

@@

\begin{flushleft}{
	\Large
	\textbf\newline{
		Inferring the generation-interval distribution that correctly links speed and strength of an epidemic
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,2,*}
David Champredon\textsuperscript{3,4}
Jonathan Dushoff\textsuperscript{3,5}
\\

@@

Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
Generation intervals are often measured through contact tracing by identifying who infected whom.
We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including truncating (due to observation time), and the effects of susceptible depletion at various spatial scales. 
Early in an epidemic, we expect the variation in the observed generation intervals to be mainly driven by the truncating and the population structure near the source of disease spread; 
therefore, we predict that correcting observed intervals for the effect of temporal truncation but \emph{not} for spatial effects will provide a spatially informed ``effective'' generation-interval distribution, which will correctly link $r$ and $\RR$.
We develop and test statistical methods for temporal corrections of generation intervals,
and confirm our prediction using individual-based simulations on an empirical network.

@@

Infectious disease modeling, generation interval, basic reproductive number, population structure, contact tracing

@@

An epidemic can be described by the exponential growth rate, $r$, and the reproductive number, \RR.
The reproductive number is defined as the average number of secondary cases arising from a primary case;
its value in a fully susceptible population, also known as the basic reproductive number \Ro, is of particular interest as it provides information about the final size of an epidemic \citep{anderson1991infectious, diekmann1990definition} as well as the endemicity level \citep{kribs2000simple, van2002reproduction, smith2012ross}.
However, estimating the reproductive number directly from disease life history requires detailed knowledge, which is not often available, particularly early in an outbreak \citep{dietz1993estimation}.
Instead, the reproductive number can be indirectly estimated from the exponential growth rate, which can be estimated from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}.
These two quantities are linked by generation-interval distributions \citep{wearing2005appropriate, svensson2007note, roberts2007model, wallinga2007generation, park2019practical}.

At the individual level, a generation \emph{interval} is defined as the time between when a person becomes infected and when that person infects another person \citep{svensson2007note}.
While this definition is widely used in the literature, it is not connected directly to a population-level \emph{distribution}.
There are important distinctions to be made when defining generation-interval distributions at the population-level. The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.
On the other hand, the \emph{realized} generation-interval distributions describe the time between actual infection events over the course of an epidemic.
Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.

The shape of the realized generation-interval distribution depends on the reference time and perspective \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic, britton2019estimation}.
When an epidemic is growing exponentially, as often occurs near the beginning of an outbreak, the number of newly infected individuals will be large relative to the number infected earlier on. 
A susceptible individual is thus relatively more likely to be infected by a newly infected individual. 
Thus, ``backward'' generation intervals (which look at a cohort of infectees, and ask when their infectors were infected) will be shorter on average than intrinsic generation intervals -- the converse is true when an epidemic is subsiding \citep{nishiura2010time, champredon2015intrinsic, britton2019estimation}.
Likewise, we can define ``forward'' generation intervals, which look at a cohort of infectors and ask when their infectees were infected.
Mean forward generation intervals can decrease over the course of an epidemic due to susceptible depletion \citep{kenah2008generation, nishiura2010time, tomba2010some, champredon2015intrinsic}.

Realized generation intervals are also affected by spatial structure.
In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population.
This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
As a result, realized generation intervals (from the perspective of an infector) will typically have shorter mean than the intrinsic generation-interval distribution in a non-homogenous population.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

While an epidemic is ongoing, realized generation intervals until the current epidemic time can be aggregated to form a single distribution.
While realized generation intervals often cannot be observed in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
we typically want to try to make inference based on all available data.
These aggregated measurements are ``truncated'' because we do not know what happens after the current epidemic time.
These truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded by the current epidemic time.
These observations allow us to make the following prediction: removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which is ``spatially corrected'' (i.e., contains information about the population structure) and will allow us to correctly infer the initial reproductive number from the exponential growth rate.

In this study, we explore spatiotemporal variation in realized (particularly forward) generation intervals.
We extend previous frameworks to study how the aggregated generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
We classify spatial effects on realized generation intervals into three levels (egocentric, local, and global) and discuss how they affect realize generation-interval distributions.
Finally, we compare two methods for accounting for temporal bias and test our prediction using simulations.

@@

Generation-interval distributions are often considered as population averages, but we can distinguish population-level distributions from individual-level distributions \citep{svensson2007note, svensson2015influence}; 
making this distinction clear will be particularly useful when we discuss spatial components later (see \fref{indpop}).
An individual-level intrinsic infection kernel $k(\tau; a)$ describes the rate at which an infected individual with ``aspect'' $a$ makes ``infectious contacts'' (contacts which will cause infection if the contactee is susceptible).
Individual aspects may represent variation in the course of infection (e.g., duration of latent and infectious periods) and the level of infectiousness, which can depend both on biological infectiousness as well as contact patterns.
Hereafter, we use $t, s$ to represent calendar time and $\tau$ to
represent time since infection.

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations:
\begin{equation}
K(\tau) = \int k (\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
The population-level kernel describes the rate at which infectious contacts are made by an an infected individual, on \emph{average}.

Assuming that a population mixes homogeneously, we can write: 
\begin{equation}
K(\tau) = \RR_0 g(\tau),
\end{equation}
where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population \citep{anderson1991infectious}), 
and $g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
Since all infectious contacts lead to infection in a disease-free equilibrium of a homogeneously mixing population, $g(\tau)$ also describes the expected time distribution of secondary cases caused by an average primary case in a fully susceptible population and can be treated as an ``equilibrium'' distribution of the corresponding homogeneous mixing model.
